A Collection of Thoughts on the Keccak/SHA-3 Construction

### 2018-05-07 s2018-math490/thesis/thesis.tex
Updates
Signed-off-by: Alexander Scheel <alexander.m.scheel@gmail.com>

    $\iota$ is a fixed-value XOR useful for preventing trivial fixed points,
and is the only function dependent on the current round number. Refer to
page 16 of FIPS 202 (\cite{NIST202}) for its specification; it is not discussed
extensively in this paper.

    The sponge construction can be described as follows. Fix $w$ as a power of
two. To have a security margin of $m$ bits, the output needs to have $m$ bits;
however, if the attacker has complete control of all $25w$ bits of state,
they would be able to choose the output easily because Keccak is bijective.
Thus, restrict the input to $25w - 2m$ bits and set the remaining $2m$ bits to
zero. To find a preimage for a $m$-bit output, the attacker would have to find
a $25w - m$ bit suffix such that the preimage of that suffix has $2m$ zeros.
Similarly, for a collision attack, an attacker has to find two $25w-2m$
messages such that their hashes have the same prefix of length $m$. Under an
ideal round permutation, clearly these are non-trivial and hard to find in
general.

    For a multiple block message the procedure is thus as follows: call the
composition of 24 rounds to be \texttt{KECCAK-$f$}. Start with $s=0^{25w}$.
Splitting the message into chunks of $25w-2m$ (and padding according to
the scheme defined in FIPS 202 \cite{NIST202}), for each block $b$, extend
$b$ to length $25w$ with zeros, then compute $s = f(s \oplus b)$. From $s$,
take the top $m$ bits as the output of the hash function.

@@

This follows from the definition of $\theta$: note that
$\theta(0^{25w}) = 0^{25w}$. If $a \neq 0^{25w}$, then there
exists an index set $I_{1}$ such that $\forall i \in I_{1}$, $a[i] = 1$,
and $\forall j \not\in I_{1}$, $a[j] = 0$. Then $\theta(a) \neq 0^{25w}$,
which follows from the construction of $\theta$. (Note that each output
bit of $\theta$ is composed of the XOR of 11 values in a fixed pattern).

@@

This follows from the definition of $\theta$: note that $\theta$ is
composed entirely of XORs and that XOR is commutative and associative.

@@

To construct the inverse of $\theta$, note that
$A'[x, y, z] = A[x, y, z] \oplus D[x, z]$; hence,
$A[x, y, z] = A'[x, y, z] \oplus D'[x, z]$ for some $D' = D$. Since $D[x, z]$
is composed of several $C[x, z]$, where
$C[x, z] = \bigoplus_{y = 0}^{4} A[x, y, z]$, we can similarly define $C'[x, z]$
to be $C'[x, z] = \bigoplus_{y = 0}^{4} A'[x, y, z]$. Then, we expect the inverse
of $\theta$ to be of a similar form. This reduces to a linear algebra problem
over boolean variables. We know that $D'[x, z] = D[x, z]$ in order to recover
$A[x, y, z]$. Hence we can represent $D[x, z]$ as strings of length
$5 \times w$, where bit $i = z' + 5 \times x'$ is $1$ if and only if
$C[x', z']$ is used in the construction of $D[x, z]$. Further, we can view each
of the $C'[x, z]$ as being the conjunction of three $C[x', z']$ in $A$; thus
these are strings where bit $j = z' + 5\times x'$ is 1 if and only if
$C[x', z']$ is used to construct $C'[x, z]$.

(That is, since $C'[x, z] = \bigoplus_{y = 0}^{4} A'[x, y, z]$,
$C'[x, z] = \bigoplus_{y = 0}^{4} (A[x, y, z] \oplus D[x, z])$ and hence
$C'[x, z] = \bigoplus_{y = 0}^{4} (A[x, y, z] \oplus C[x', z'] \oplus C[x'', z''])$,
and thus $C'[x, z] = C[x, z] \oplus C[x', z'] \oplus C[x'', z'']$, for some
$x, z, x', z', x'', z''$ based on the definition of $\theta$).

Thus, giving each $C'[x, z]$ a constant $c_{x,z}$ for whether it is used in
constructing $D'[x, z]$, we can form a system of linear equations and solve for
the constants $c_{x, z}$ in each expression. Since there are $5 \times w$
variables and $5 \times w$ equations in each equation for $D'[x, z]$, this can
be solved easily, yielding the inverse of $\theta$.

>>>>

    $\iota$ is a fixed-value XOR useful for preventing trivial fixed points,
and is the only function dependent on the current round number. Refer to
page 16 of FIPS 202 (\cite{NIST202}) for its specification; it is not discussed
extensively in this paper. It takes an additional parameter, $c$, the current
round number.

    The a single round of the core round function function, $r$ is defined as
follows:
\begin{align*}
    r(A, c) = \iota(\chi(\pi(\rho(\theta(A)))), c)
\end{align*}
And the entire core round function, \texttt{KECCAK-$f$}, is thus:
\begin{align*}
    \text{\texttt{KECCAK-$f$}}(A) = r(r( ...  r(A, 0) ... , 22), 23)
\end{align*}

    The sponge construction can be described as follows. Fix $w$ as a power of
two in $W$. Then define a state, $s$, as above to be a binary string of length
$25w$. Fix a security margin $m$ (usually 228, 256, 384, or 512 bits in the
SHA-3 standard). Define $i_{m} = 25w - 2m$ be the input margin, and
$o_{m} = 2m$ be the output margin. For a message $M$, assume it is padded
to a multiple of $i_{m}$ bits according to the scheme in FIPS 202
(\texttt{pad10*1}). Start with initial state $s_0 = 0^{25w}$. For each block
$M_{i}$, XOR $M_{i}$ into the first $|M_{i}|$ bits of state $s_i$ and define:
\begin{align*}
    s_{i+1} = \text{\texttt{KECCAK-$f$}}(s_{i} \oplus M_{i})
\end{align*}
Then if $s_{k}$ is the final state, take the first $m$ bits ($s_{k}[0..m]$)
as the output digest. Additionally if more than $2m$ bits are required, take
the first $2m$ bits, then compute $s_{k+1} = \text{\texttt{KECCAK-$f$}}(s_{k})$
and continue taking at most the first $2m$ bits from $s_{k+1}$, repeating as
necessary. This is demonstrated in Figure \ref{f:s}.

\begin{figure*}
\begin{center}
\includegraphics[width=0.7\textwidth]{figs/Sponge-150.png}
\caption{Graphical description of the sponge construction. Courtesy of the
\href{https://keccak.team/images/Sponge-150.png}{Keccak Team}.}
\label{f:s}
\end{center}
\end{figure*}

To have a security margin of $m$ bits, the output needs to have $m$ bits;
however, if the attacker has complete control of all $25w$ bits of state,
they would be able to choose an output with prefix $m$ easily because
Keccak is bijective with a known, computable inverse. Thus, by restricting
the initial state to have a suffix of $2m$ bits of zeros, the attacker must
not only restrict the output to have the correct prefix, but also the input
to have the correct suffix. In this way, the authors claim that Keccak/SHA-3
has collision resistance equal to preimage resistance; finding a collision
is of equivalent difficulty as finding a preimage, and thus Keccak/SHA-3 is
more secure than necessary \cite{Keccak3}.

Further, SHA-3 is secure from trivial multiple-block collisions due to the
restriction of allowing $i_{m}$-bits of input state to be changed directly
via a block. Thus, for a multiple-block collision attack to be possible, the
remaining $25w - i_{m} = 2m$ bits must be the same between the internal state
of a collision pair, allowing two different blocks $b$ and $b'$ to cancel
their initial prefix differences. All of these attacks are constrained
input/output problems and thus reduce to SAT; thus, considerable work
is required to successfully perform an attack.

@@

First, note that $\theta(0^{25w}) = 0^{25w}$ due to construction of $\theta$
having an even number of XORs. Suppose that there existed some other input, x,
to $\theta$ such that $\theta(x) = 0^{25w}$. Note that, $x$ with odd number of
1s (that is, $|\{ i : x_{i} = 1 \}| \equiv 1 \text { mod } 2$), $\theta(x)$
likewise has an odd number of ones due to $\theta$ having an even number of
XORs. Similarly, for any $x$ with has an even but non-zero number of ones,
then there exists index $(i, j, k)$ such that $\theta(x)[i, j, k] = 1$. This is
because, for any pair of indices $(i, j, k)$ and $(i', j', k')$ such that
$x[i, j, k] = 1 = x[i', j', k']$, there exists an index $(a, b, c)$ such that
$\theta(x)[a, b, c]$ is dependent on only one of $(i, j, k)$ or $(i', j', k')$,
by construction of $\theta$. Thus the kernel of $\theta is 0^{25w}$.

@@

This follows from the definition of $\theta$: note that $\theta$ is
composed entirely of XORs and that XOR is commutative and associative. For any
index $(i, j, k)$, we want to show that
\begin{align*}
    \theta(a \oplus b)[i, j, k] = \theta(a)[i, j, k] \oplus \theta(b)[i, j, k]
\end{align*}
Note that:
\begin{align*}
    \theta(a \oplus b)[i, j, k] & = (a[i, j, k] \oplus b[i, j, k]) \\
        & \oplus (\bigoplus_{0 \leq y \leq 4} (a[(i - 1) \text{ mod } 5, j, k] \oplus b[(i - 1) \text{ mod } 5, j, k])) \\
        & \oplus (\bigoplus_{0 \leq y \leq 4} (a[(i + 1) \text{ mod } 5, j, (k - 1) \text{ mod } w] \oplus b[(i + 1) \text{ mod } 5, j, (k - 1) \text{ mod } w])) \\
        & = (a[i, j, k ] \oplus (\bigoplus_{0 \leq y \leq 4} a[(i - 1) \text{ mod } 5, j, k]) \oplus (\bigoplus_{0 \leq y \leq 4} a[(i + 1) \text{ mod } 5, j, (k - 1) \text{ mod } w])) \\
        & \oplus (b[i, j, k ] \oplus (\bigoplus_{0 \leq y \leq 4} b[(i - 1) \text{ mod } 5, j, k]) \oplus (\bigoplus_{0 \leq y \leq 4} b[(i + 1) \text{ mod } 5, j, (k - 1) \text{ mod } w])) \\
        & = \theta(a)[i, j, k] \oplus \theta(b)[i, j, k]
\end{align*}
\end{proof}

@@

To construct the inverse of $\theta$, note that since $\theta$ is a linear map,
there exists a matrix $M$ such that $\theta(x) = Mx$, when $x$ is treated as a
column vector of bits. $M$ is described by a square matrix of size $25w$ by
$25w$, where for a given row $i$:
\begin{align*}
    M_{i, j} = 1 \iff \theta(x)[i] \text{ depends on } x[j]
\end{align*}
Since we know that $\theta$ is invertible, $M$ must be invertible and thus
$M^{-1}$ describes the inverse of $\theta$.
